---
title: What is an ROC-curve and why is it concave?
output: pdf_document
date: '2020-08-18'
tags:
- statistics 
permalink: '`r paste0("/posts/",sub("[.]Rmd","",knitr::current_input()),"/")`'
postname: '`r sub("[.]Rmd","",knitr::current_input())`'
pdf: true
rmd: true
header-includes:
  - \usepackage{mathrsfs}
---

Probably the most ubiquitous evaluation metric for prediction models for binary outcomes (be it diagnostic or prognostic), is the AUC, also named C-statistic, or in full: the area under the curve of the receiver operator curve.
Recently I came across an ROC-curve that seemed a bit odd to me. It was published [here](doi:10.1001/jamanetworkopen.2019.6879) (doi:10.1001/jamanetworkopen.2019.6879) in JAMA Network Open and included an ROC curve that curled up on itself. 

https://cdn.jamanetwork.com/ama/content_public/journal/jamanetworkopen/938081/zoi190277f3.png?Expires=2147483647&Signature=C~Gn712RBY43Xlq3WX1OMdxspU60uTptbeDkOCWp58WWn-zTQz6UCiGsSLLwI10hwFxpZYScTW-fxiXLUpiKwoFtpJkIGMuyqSBgYB5Ewt4K~9bSxGcVIG1vBiclRZq2a9jQ0rjoOnI2aNNBxal0hJCaDuWHWf1q9vh1Gi2osC3kaEXlWX4K1jxX0Z3lOUpinLSt-6oMkhsJqWTPYO2Ud0PLvuQayOC8psmj3STQNba9VTbCZoVPQf77zV3qm0KFC30yd6I1NyjWBnFkv7-I9XvF00kHebkKGNeAx9dFOGbIzondj0XFVRL8yOi8FlO~8hUkEGoscc8vBEXX~8IU0A__&Key-Pair-Id=APKAIE5G5CRDK6RD3PGA

I wrote this blog-post to see if this is at all possible, or whether it is a sign of some error.
In this post I will explain what the ROC-curve is and how you can 'manually' construct one in R, using only sorting rules and simple math.

### The basis of the ROC-curve: Sensitivity and Specificity

Formally, an ROC-curve is the visualization of the sensitivity and specificity of a continuous predictor for predicting a binary outcome, using different cut-offs.
The most common example in medical literature is using logistic regression to predict a binary outcome, e.g. benign vs malignant breast tumours.




To see whether this is at all possible, or whether this is a sign of some error, I wrote 

To see this is possible, we will use data on smoking, alcohol and esophageal cancer

```{r, imports}
library(data.table)
library(pROC)
data(esoph)
```

```{r}
df <- MASS::biopsy
```

```{r}
xycols <- c("V1", "V2", "V3", "V4", "V5", "V6", "V7", "V8", "V9", "class")
m1 <- glm(class~., data=df[, xycols], family=binomial)

thickness        <- df$V1
size_uniformity  <- df$V2
shape_uniformity <- df$V3
malignant        <- df$class == 'malignant'

m1 <- glm(malignant ~ thickness + size_uniformity + shape_uniformity, family=binomial)

predictions <- predict(m1, type='response')
```

```{r}
prediction_order    <- order(predictions)
malignant_ordered   <- malignant[prediction_order]
predictions_ordered <- predictions[prediction_order]

sens <- function(y_true, y_pred) sum(y_pred[y_true]) / sum(y_true)
spec <- function(y_true, y_pred) sum(!y_pred[!y_true]) / sum(!y_true)

unique_prediction_values <- unique(predictions_ordered)
n_unique_prediction_values <- length(unique_prediction_values)

cutoffs       <- vector(mode='numeric', length=n_unique_prediction_values)
sensitivities <- vector(mode='numeric', length=n_unique_prediction_values)
specificities <- vector(mode='numeric', length=n_unique_prediction_values)

for (i in 1:n_unique_prediction_values) {
  cutoff            <- unique_prediction_values[i]
  cutoffs[i]        <- cutoff
  binary_prediction <- predictions_ordered > cutoff
  sensitivities[i]  <- sens(malignant_ordered, binary_prediction)
  specificities[i]  <- spec(malignant_ordered, binary_prediction)
}

```

```{r}
plot(1 - specificities, sensitivities, type='l')
abline(a=0, b=1)
```

```{r}
proc <- pROC::roc(malignant~predictions)
plot(proc)
```






https://www.kaggle.com/sulianova/cardiovascular-disease-dataset/data#

