---
title: What is an ROC-plot and why is it monotonic?
output: pdf_document
date: '2020-08-18'
tags:
- statistics 
permalink: '`r paste0("/posts/",sub("[.]Rmd","",knitr::current_input()),"/")`'
postname: '`r sub("[.]Rmd","",knitr::current_input())`'
pdf: true
rmd: true
header-includes:
  - \usepackage{mathrsfs}
---

Probably the most ubiquitous evaluation metric for prediction models for binary outcomes (be it diagnostic or prognostic), is the AUC, also named C-statistic, or in full: the area under the curve of the receiver operator curve.
Recently I came across an ROC-curve that seemed a bit odd to me. It was published [here](doi:10.1001/jamanetworkopen.2019.6879) (doi:10.1001/jamanetworkopen.2019.6879) in JAMA Network Open and included an ROC curve that curled up on itself. 

https://cdn.jamanetwork.com/ama/content_public/journal/jamanetworkopen/938081/zoi190277f3.png?Expires=2147483647&Signature=C~Gn712RBY43Xlq3WX1OMdxspU60uTptbeDkOCWp58WWn-zTQz6UCiGsSLLwI10hwFxpZYScTW-fxiXLUpiKwoFtpJkIGMuyqSBgYB5Ewt4K~9bSxGcVIG1vBiclRZq2a9jQ0rjoOnI2aNNBxal0hJCaDuWHWf1q9vh1Gi2osC3kaEXlWX4K1jxX0Z3lOUpinLSt-6oMkhsJqWTPYO2Ud0PLvuQayOC8psmj3STQNba9VTbCZoVPQf77zV3qm0KFC30yd6I1NyjWBnFkv7-I9XvF00kHebkKGNeAx9dFOGbIzondj0XFVRL8yOi8FlO~8hUkEGoscc8vBEXX~8IU0A__&Key-Pair-Id=APKAIE5G5CRDK6RD3PGA

To see this is possible, we will use data on smoking, alcohol and esophageal cancer

```{r, imports}
library(data.table)
library(pROC)
data(esoph)
```

```{r}
m1 <- glm()
```

