---
title: Finding the functional form for multiple linear regression
output: pdf_document
date: '2019/08/16'
date-modified: "2024/03/22"
categories:
- statistics 
- simulations
- linear regression
---

## Functional relationship in linear regression

A frequent question that comes up when modeling continuous outcomes with multiple linear regression 
is what the correct functional form for the relationship between the independent variables is.
TLDR: the answer is a [partial residual plot](https://en.wikipedia.org/wiki/Partial_residual_plot). 
Here I will generate some data to illustrate this

## Data


```{r}
suppressMessages({require(ggplot2); theme_set(theme_bw())})
set.seed(12345)
N = 1000
x <- runif(N, min = 0, max=2*pi)
w <- .5*x + sin(x) + rnorm(N, sd=.25)
sy <- rnorm(N, sd=.1)
y <- x + w + sy
df = data.frame(x,w,y)
```


The data consists of two real-valued 'independent' variables $x,w$, 
where 

$$
\begin{align}
x &\sim U(0, 2 \pi) \\
\epsilon_w &\sim N(0, 0.25) \\
\epsilon_y &\sim N(0, 0.1) \\
w &:= \frac{x}{2} + \sin(x) \\
y &:= x + w + \epsilon_y
\end{align}
$$

Clearly $y$ is linear in both $x$ and $w$.
A plot of the data:

```{r}
#| fig-cap: "joint distribution of x, w, y"
ggplot(df, aes(x=x, y=w,col=y, size=y)) + 
  geom_point() + theme_minimal()
```

## Plots

### Incorrect: marginal association


Let's say we're particularly interested in the relationship between $y$ and $x$, both conditional on $w$.
Looking at the *marginal* association between $y$ and $x$ with a scatterplot will set us on the wrong foot, 
because of the association between $x$ and $w$.

```{r}
#| fig-cap: "marginal association between x and y"
ggplot(df, aes(x=x,y=y)) + 
  geom_point()
```

### Correct: partial residual plot

To construct the correct plot, we can generate a partial residual plot, which is created with `resid(lm(y~x+w))+b_x x ~ x`.

Where `b_x` is the regression coefficient found through linear regression of $y$ on $x$ and $w$.
In a plot:

```{r}
#| label: fig-partialresid
#| fig-cap: "partial residual plot"
lmfit <- lm(y~x+w)
yresid <- resid(lmfit)
b_x <- coef(lmfit)['x']
plotdata <- data.frame(x, y=yresid + b_x * x)

ggplot(plotdata, aes(x=x,y=y)) + 
  geom_point() + 
  ylab("resid(lm(y~x+w)) + b_x x")
```

### Partial residual plot when y is not linear in x

What if $y$ were not linear in $x$?

```{r}
#| fig-cap: "partial residual plot when y not linear in x"
y2 <- x^2 + w + sy
lmfit2 <- lm(y2 ~ x + w)
b_x2 <- coef(lmfit2)['x']
plotdata2 <- data.frame(x, y=resid(lmfit2) + b_x2 * x)

ggplot(plotdata2, aes(x=x,y=y)) + 
  geom_point() + 
  ylab("resid(lm(y~x+w)) + b_x x")

```

## Conclusion

The functional relationship between an outcome and a covariate in a linear regression conditional on other covariates is visualized with a [partial residual plot](https://en.wikipedia.org/wiki/Partial_residual_plot), or:

```{r}
#| eval: false
lmfit <- lm(y~x+w)
plot(resid(lmfit) + coef(lmfit)['x'] * x ~ x)
```



